---
title: "cTWAS Results Summary-AD GWAS (Bellenguez et al. 2022)"
author: "XSun"
date: "2026-06-19"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ctwas)
library(ggplot2)
library(DT)
library(ggrepel)
library(dplyr)
library(RSQLite)
library(VennDiagram)
library(grid)
library(EnsDb.Hsapiens.v86)

gwas_n <- 487511
outputdir <- "/project/xinhe/xsun/AD_jubao/3.ctwas/output"
qtl_types <- c("caQTL", "eQTL")

plot_piechart_topn <- function(ctwas_parameters, colors, by, title, n_tissue = 5) {
  data <- data.frame(
    category = names(ctwas_parameters$prop_heritability),
    percentage = ctwas_parameters$prop_heritability
  )

  data <- data %>%
    mutate(
      context = sub("\\|.*", "", category),
      type = sub(".*\\|", "", category)
    )

  if (by == "type") {
    data <- data %>%
      group_by(type) %>%
      summarize(percentage = sum(percentage)) %>%
      mutate(category = type)
  } else if (by == "context") {
    data <- data %>%
      group_by(context) %>%
      summarize(percentage = sum(percentage)) %>%
      mutate(category = context)
  } else if (by == "category") {
    data <- data %>%
      mutate(category = ifelse(category == "SNP", "SNP", category))
  } else {
    stop("Invalid 'by' parameter. Use 'type', 'context', or 'category'.")
  }

  data$percentage_label <- paste0(round(data$percentage * 100, 2), "%")

  if (nrow(data) > (n_tissue + 1)) {
    data <- data %>%
      filter(context != "SNP") %>%
      arrange(desc(percentage)) %>%
      mutate(rank = row_number()) %>%
      mutate(context = ifelse(rank <= 5, context, "Other_Tissues")) %>%
      group_by(context) %>%
      summarise(percentage = sum(percentage), .groups = "drop") %>%
      bind_rows(data %>% filter(context == "SNP") %>% select(context, percentage)) %>%
      mutate(
        category = context,
        percentage_label = paste0(sprintf("%.2f", percentage * 100), "%")
      ) %>%
      arrange(desc(percentage))

    sorted_levels <- data %>%
      mutate(sort_key = case_when(
        category == "Other_Tissues" ~ 1,
        category == "SNP" ~ 2,
        TRUE ~ 0
      )) %>%
      arrange(sort_key, desc(percentage)) %>%
      pull(category)
  } else {
    sorted_levels <- data %>%
      arrange((category == "SNP"), desc(percentage)) %>%
      pull(category)
  }

  data$category <- factor(data$category, levels = sorted_levels)

  data <- data %>%
    arrange(category) %>%
    mutate(
      cumulative = cumsum(percentage),
      midpoint = cumulative - percentage / 2
    )

  categories <- levels(data$category)
  has_snp <- "SNP" %in% categories
  other_cats <- setdiff(categories, "SNP")

  color_vec <- c()
  if (has_snp) {
    color_vec["SNP"] <- "#fbf6ef"
  }

  if (length(other_cats) > 0) {
    if (is.null(names(colors))) {
      colors_other <- rep(colors, length.out = length(other_cats))
      color_vec <- c(color_vec, setNames(colors_other, other_cats))
    } else {
      for (cat in other_cats) {
        color_vec[cat] <- ifelse(cat %in% names(colors), colors[cat], "#c9cba3")
      }
    }
  }

  data <- data %>%
    mutate(
      y_pos = midpoint,
      angle = 0,
      left_side = midpoint > 0.5,
      hjust = ifelse(left_side, 1, 0)
    )

  pie <- ggplot(data, aes(x = "", y = percentage, fill = category)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    theme_void() +
    geom_text_repel(
      aes(y = 1 - y_pos, label = percentage_label, angle = angle),
      size = 3,
      nudge_x = 0.8,
      segment.size = 0.3,
      segment.color = "gray40",
      box.padding = 0.2,
      min.segment.length = 0.1,
      hjust = 0.5,
      vjust = 0.5
    ) +
    scale_fill_manual(values = color_vec) +
    labs(fill = "") +
    ggtitle(title)

  return(pie)
}
```


# Analysis details

The cTWAS analysis was performed using `run_ctwas.R` located at `/project/xinhe/xsun/AD_jubao/3.ctwas/run_ctwas.R`.

**Key steps:**

1. **GWAS preprocessing**: GWAS summary statistics from [Bellenguez et al. 2022](https://pubmed.ncbi.nlm.nih.gov/35379992/) were read and preprocessed with `read_gwas()` and `preprocess_z_snp()` (dropping multi-allelic and strand-ambiguous variants).
2. **Weight preprocessing**: Prediction weights (PredictDB format, RASQUAL-based) were preprocessed with `preprocess_weights()` for each condition (naive, SL1344). For eQTL, protein-coding gene filtering was applied; for caQTL it was not.
3. **Gene z-score computation**: Gene-level z-scores were computed with `compute_gene_z()`.
4. **Region data assembly**: Region data was assembled with `assemble_region_data()` using `thin = 1` and `adjust_boundary_genes = TRUE`.
5. **Parameter estimation**: Parameters were estimated with `est_param()` using `group_prior_var_structure = "shared_all"`, `null_method = "ctwas"`, `niter_prefit = 3`, `niter = 30`, `min_gene = 1`, `min_var = 2`, `min_p_single_effect = 0.8`.
6. **Fine-mapping**: Fine-mapping was performed with `finemap_regions()` using `L = 5` and `null_method = "ctwas"`.

**LD reference**: UK Biobank LD reference (b38) stored at `/project2/compbio/UKB_LD_REF/UKB_LDR_0.1_b38/`.

**Weight files** (RASQUAL-based):

- caQTL: `caQTL_naive_RASQUAL.db`, `caQTL_SL1344_RASQUAL.db`
- eQTL: `eQTL_naive_RASQUAL.db`, `eQTL_SL1344_RASQUAL.db`

# Molecular traits used in weights

## caQTL

We compare the ATAC peaks present in the naive and SL1344 caQTL weight databases.

```{r caQTL-peaks, fig.width=6, fig.height=5}
weight_dir <- "/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/RASQUAL/weights_rasqual"

db_naive <- dbConnect(SQLite(), file.path(weight_dir, "caQTL_naive_RASQUAL.db"))
peaks_naive <- dbGetQuery(db_naive, "SELECT gene FROM extra")$gene
dbDisconnect(db_naive)

db_sl <- dbConnect(SQLite(), file.path(weight_dir, "caQTL_SL1344_RASQUAL.db"))
peaks_sl <- dbGetQuery(db_sl, "SELECT gene FROM extra")$gene
dbDisconnect(db_sl)

n_naive <- length(peaks_naive)
n_sl <- length(peaks_sl)
n_shared <- length(intersect(peaks_naive, peaks_sl))
n_total <- length(union(peaks_naive, peaks_sl))

grid.newpage()
draw.pairwise.venn(
  area1 = n_naive,
  area2 = n_sl,
  cross.area = n_shared,
  category = c("Naive", "SL1344"),
  scaled = TRUE,
  fill = c("#E69F00", "#56B4E9"),
  alpha = 0.5,
  cat.pos = c(-20, 20),
  cat.dist = 0.05,
  cat.cex = 1.2,
  cex = 1.1,
  label.col = "black",
  ext.text = FALSE,
  print.mode = c("raw", "percent")
)
grid.text("caQTL: Peaks in Weight Databases", y = 0.95, gp = gpar(fontsize = 14))
```

## eQTL

We compare the genes present in the naive and SL1344 eQTL weight databases.

```{r eQTL-genes, fig.width=6, fig.height=5}
db_naive_e <- dbConnect(SQLite(), file.path(weight_dir, "eQTL_naive_RASQUAL.db"))
genes_naive <- dbGetQuery(db_naive_e, "SELECT gene FROM extra")$gene
dbDisconnect(db_naive_e)

db_sl_e <- dbConnect(SQLite(), file.path(weight_dir, "eQTL_SL1344_RASQUAL.db"))
genes_sl <- dbGetQuery(db_sl_e, "SELECT gene FROM extra")$gene
dbDisconnect(db_sl_e)

n_naive_e <- length(genes_naive)
n_sl_e <- length(genes_sl)
n_shared_e <- length(intersect(genes_naive, genes_sl))

grid.newpage()
draw.pairwise.venn(
  area1 = n_naive_e,
  area2 = n_sl_e,
  cross.area = n_shared_e,
  category = c("Naive", "SL1344"),
  scaled = TRUE,
  fill = c("#E69F00", "#56B4E9"),
  alpha = 0.5,
  cat.pos = c(-20, 20),
  cat.dist = 0.05,
  cat.cex = 1.2,
  cex = 1.1,
  label.col = "black",
  ext.text = FALSE,
  print.mode = c("raw", "percent")
)
grid.text("eQTL: Genes in Weight Databases", y = 0.95, gp = gpar(fontsize = 14))
```

# cTWAS results

## caQTL

```{r caQTL-load}
param_caQTL <- readRDS(file.path(outputdir, "caQTL", "AD.caQTL.param.shared_all.RDS"))
```

### Convergence

```{r caQTL-convergence, fig.width=10, fig.height=6}
make_convergence_plots(param_caQTL, gwas_n)
```

### Parameters and Enrichment

```{r caQTL-param}
ctwas_param_caQTL <- summarize_param(param_caQTL, gwas_n, enrichment_test = "fisher")
ctwas_param_caQTL
```

### Heritability Pie Chart

```{r caQTL-piechart, fig.width=7, fig.height=5}
plot_piechart_topn(ctwas_param_caQTL,
                   colors = c("#E69F00", "#56B4E9"),
                   by = "category",
                   title = "caQTL - Percent of Heritability (PVE)")
```

### Fine-mapping results (PIP > 0.8)

```{r caQTL-finemap}
finemap_caQTL <- readRDS(file.path(outputdir, "caQTL",
                         "AD.caQTL.shared_all.L5.finemap_regions_res.RDS"))
fm_caQTL <- finemap_caQTL$finemap_res

peaks_high_pip <- fm_caQTL[fm_caQTL$type != "SNP" & fm_caQTL$susie_pip > 0.8, ]
peaks_high_pip <- peaks_high_pip[order(-peaks_high_pip$susie_pip), ]
```

**`r nrow(peaks_high_pip)` peaks** with PIP > 0.8:

```{r caQTL-finemap-table}
if (nrow(peaks_high_pip) > 0) {
  datatable(peaks_high_pip,
            caption = "caQTL peaks with PIP > 0.8",
            options = list(pageLength = 20, scrollX = TRUE))
} else {
  cat("No peaks found with PIP > 0.8.")
}
```

### Locus zoom plots

Locus zoom plots for regions containing ATAC peaks with PIP > 0.8.

```{r caQTL-locusplot-prep}
ens_db <- EnsDb.Hsapiens.v86
weights_caQTL <- readRDS(file.path(outputdir, "caQTL", "AD.caQTL.preprocessed.weights.RDS"))
snp_map <- readRDS("/project2/xinhe/shared_data/multigroup_ctwas/LD_region_info/snp_map.RDS")
mapping_table_caQTL <- readRDS("/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/ATAC_peak_mapping.RDS")

fm_caQTL_anno <- fm_caQTL
fm_caQTL_anno$pval <- z2p(fm_caQTL_anno$z)

fm_caQTL_anno <- anno_finemap_res(fm_caQTL_anno,
                                  snp_map = snp_map,
                                  mapping_table = mapping_table_caQTL,
                                  add_gene_annot = TRUE,
                                  map_by = "molecular_id",
                                  drop_unmapped = TRUE,
                                  add_position = TRUE,
                                  use_gene_pos = "mid")

top_regions_caQTL <- unique(fm_caQTL_anno$region_id[fm_caQTL_anno$group != "SNP" & fm_caQTL_anno$susie_pip > 0.8])
```

```{r caQTL-locusplots, fig.width=10, fig.height=8, results='asis'}
for (region_id in top_regions_caQTL) {
  cat(sprintf("\n#### Region: %s\n\n", region_id))

  print(make_locusplot(
    fm_caQTL_anno,
    region_id = region_id,
    ens_db = ens_db,
    weights = weights_caQTL,
    highlight_pip = 0.8,
    filter_protein_coding_genes = FALSE,
    filter_cs = F,
    color_pval_by = "cs",
    color_pip_by = "cs",
    panel.heights = c(4, 4, 1, 4)
  ))

  cat("\n\n")
}
```

### Peak Sets

A **peak set** is defined as all caQTL peaks that belong to the same credible set (CS) within the same LD region. The PIP of a peak set is the sum of the individual peak PIPs.

```{r caQTL-peak-sets}
peaks_in_cs_caQTL <- fm_caQTL[fm_caQTL$type == "caQTL" & !is.na(fm_caQTL$cs), ]

mapping_caQTL <- readRDS("/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/ATAC_peak_mapping.RDS")
peaks_annot_caQTL <- merge(peaks_in_cs_caQTL, mapping_caQTL, by = "molecular_id", all.x = TRUE)

peak_sets_caQTL <- do.call(rbind, lapply(
  split(peaks_annot_caQTL, list(peaks_annot_caQTL$region_id, peaks_annot_caQTL$cs), drop = TRUE),
  function(x) {
    region_parts <- strsplit(x$region_id[1], "_")[[1]]
    data.frame(
      region_chr   = as.integer(region_parts[1]),
      region_id    = x$region_id[1],
      cs           = x$cs[1],
      n_peaks      = nrow(x),
      sum_pip      = sum(x$susie_pip),
      peaks        = paste(x$molecular_id, collapse = ", "),
      contexts     = paste(x$context, collapse = ", "),
      peak_coords  = paste0("chr", x$chrom, ":", x$start, "-", x$end, collapse = "; "),
      stringsAsFactors = FALSE
    )
  }
))
peak_sets_caQTL <- peak_sets_caQTL[order(-peak_sets_caQTL$sum_pip), ]
rownames(peak_sets_caQTL) <- NULL

cat("**Total peak sets:**", nrow(peak_sets_caQTL), "\n")
```

### PIP Distribution of Peak Sets

```{r caQTL-peak-sets-pip-hist, fig.width=6, fig.height=4}
hist(peak_sets_caQTL$sum_pip, breaks = 20, col = "steelblue", border = "white",
     main = "Distribution of peak set PIPs",
     xlab = "Sum PIP per peak set", ylab = "Count")
abline(v = 0.8, col = "red", lty = 2)
text(0.8, par("usr")[4] * 0.9, "PIP = 0.8", pos = 2, col = "red", cex = 0.8)
```

### High-Confidence Peak Sets (sum PIP > 0.8)

```{r caQTL-peak-sets-high}
high_caQTL <- peak_sets_caQTL[peak_sets_caQTL$sum_pip > 0.8, ]

if (nrow(high_caQTL) > 0) {
  display_high_caQTL <- high_caQTL[, c("region_chr", "region_id", "cs", "n_peaks",
                                        "sum_pip", "peaks", "contexts", "peak_coords")]
  colnames(display_high_caQTL) <- c("Chr", "Region", "CS", "N peaks",
                                     "Sum PIP", "Peaks", "Contexts", "Peak coordinates")
  datatable(display_high_caQTL, rownames = FALSE,
            options = list(pageLength = 25, scrollX = TRUE),
            caption = "Peak sets with sum PIP > 0.8") %>%
    formatRound("Sum PIP", digits = 4)
} else {
  cat("No peak sets with sum PIP > 0.8.\n")
}
```

### Locus zoom plots for high-confidence peak sets

Locus zoom plots for regions containing peak sets with sum PIP > 0.8.

```{r caQTL-peakset-locusplot-prep}
top_regions_peakset_caQTL <- unique(high_caQTL$region_id)
```

```{r caQTL-peakset-locusplots, fig.width=10, fig.height=8, results='asis'}
for (region_id in top_regions_peakset_caQTL) {
  cat(sprintf("\n#### Region: %s\n\n", region_id))

  print(make_locusplot(
    fm_caQTL_anno,
    region_id = region_id,
    ens_db = ens_db,
    weights = weights_caQTL,
    highlight_pip = 0.8,
    filter_protein_coding_genes = FALSE,
    filter_cs = F,
    color_pval_by = "cs",
    color_pip_by = "cs",
    panel.heights = c(4, 4, 1, 4)
  ))

  cat("\n\n")
}
```






## eQTL

```{r eQTL-load}
param_eQTL <- readRDS(file.path(outputdir, "eQTL", "AD.eQTL.param.shared_all.RDS"))
```

### Convergence

```{r eQTL-convergence, fig.width=10, fig.height=6}
make_convergence_plots(param_eQTL, gwas_n)
```

### Parameters and Enrichment

```{r eQTL-param}
ctwas_param_eQTL <- summarize_param(param_eQTL, gwas_n, enrichment_test = "fisher")
ctwas_param_eQTL
```

### Heritability Pie Chart

```{r eQTL-piechart, fig.width=7, fig.height=5}
plot_piechart_topn(ctwas_param_eQTL,
                   colors = c("#E69F00", "#56B4E9"),
                   by = "category",
                   title = "eQTL - Percent of Heritability (PVE)")
```

### Fine-mapping results (PIP > 0.8)

```{r eQTL-finemap}
finemap_eQTL <- readRDS(file.path(outputdir, "eQTL",
                         "AD.eQTL.shared_all.L5.finemap_regions_res.RDS"))
fm_eQTL <- finemap_eQTL$finemap_res

genes_high_pip <- fm_eQTL[fm_eQTL$type != "SNP" & fm_eQTL$susie_pip > 0.8, ]
genes_high_pip <- genes_high_pip[order(-genes_high_pip$susie_pip), ]
```

**`r nrow(genes_high_pip)` genes** with PIP > 0.8:

```{r eQTL-finemap-table}
if (nrow(genes_high_pip) > 0) {
  datatable(genes_high_pip,
            caption = "eQTL genes with PIP > 0.8",
            options = list(pageLength = 20, scrollX = TRUE))
} else {
  cat("No genes found with PIP > 0.8.")
}
```

### Locus zoom plots

Locus zoom plots for regions containing genes with PIP > 0.8.

```{r eQTL-locusplot-prep}
weights_eQTL <- readRDS(file.path(outputdir, "eQTL", "AD.eQTL.preprocessed.weights.RDS"))
mapping_table_eQTL <- readRDS("/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/RNA_mapping.RDS")

fm_eQTL_anno <- fm_eQTL
fm_eQTL_anno$pval <- z2p(fm_eQTL_anno$z)

fm_eQTL_anno <- anno_finemap_res(fm_eQTL_anno,
                                  snp_map = snp_map,
                                  mapping_table = mapping_table_eQTL,
                                  add_gene_annot = TRUE,
                                  map_by = "molecular_id",
                                  drop_unmapped = TRUE,
                                  add_position = TRUE,
                                  use_gene_pos = "mid")

top_regions_eQTL <- unique(fm_eQTL_anno$region_id[fm_eQTL_anno$group != "SNP" & fm_eQTL_anno$susie_pip > 0.8])
```

```{r eQTL-locusplots, fig.width=10, fig.height=8, results='asis'}
for (region_id in top_regions_eQTL) {
  cat(sprintf("\n#### Region: %s\n\n", region_id))

  print(make_locusplot(
    fm_eQTL_anno,
    region_id = region_id,
    ens_db = ens_db,
    weights = weights_eQTL,
    highlight_pip = 0.8,
    filter_protein_coding_genes = TRUE,
    filter_cs = F,
    color_pval_by = "cs",
    color_pip_by = "cs",
    panel.heights = c(4, 4, 1, 4)
  ))

  cat("\n\n")
}
```

## caQTL_SL1344 + eQTL_SL1344 (Joint)

```{r joint-load}
param_joint <- readRDS(file.path(outputdir, "joint_caQTL_eQTL_SL1344", "AD.joint.param.shared_all.RDS"))
```

### Convergence

```{r joint-convergence, fig.width=10, fig.height=6}
make_convergence_plots(param_joint, gwas_n)
```

### Parameters and Enrichment

```{r joint-param}
ctwas_param_joint <- summarize_param(param_joint, gwas_n, enrichment_test = "fisher")
ctwas_param_joint
```

### Heritability Pie Chart

```{r joint-piechart, fig.width=7, fig.height=5}
plot_piechart_topn(ctwas_param_joint,
                   colors = c("#E69F00", "#56B4E9"),
                   by = "category",
                   title = "caQTL_SL1344 + eQTL_SL1344 - Percent of Heritability (PVE)")
```

### Fine-mapping results (PIP > 0.8)

```{r joint-finemap}
finemap_joint <- readRDS(file.path(outputdir, "joint_caQTL_eQTL_SL1344",
                         "AD.joint.shared_all.L5.finemap_regions_res.RDS"))
fm_joint <- finemap_joint$finemap_res

traits_high_pip_joint <- fm_joint[fm_joint$type != "SNP" & fm_joint$susie_pip > 0.8, ]
traits_high_pip_joint <- traits_high_pip_joint[order(-traits_high_pip_joint$susie_pip), ]
```

**`r nrow(traits_high_pip_joint)` molecular traits** with PIP > 0.8:

```{r joint-finemap-table}
if (nrow(traits_high_pip_joint) > 0) {
  datatable(traits_high_pip_joint,
            caption = "Joint caQTL_SL1344 + eQTL_SL1344: traits with PIP > 0.8",
            options = list(pageLength = 20, scrollX = TRUE))
} else {
  cat("No traits found with PIP > 0.8.")
}
```

### Locus zoom plots

Locus zoom plots for regions containing molecular traits with PIP > 0.8.

```{r joint-locusplot-prep}
weights_joint <- readRDS(file.path(outputdir, "joint_caQTL_eQTL_SL1344", "AD.joint.preprocessed.weights.RDS"))

mapping_table_joint <- rbind(
  readRDS("/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/ATAC_peak_mapping.RDS"),
  readRDS("/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/RNA_mapping.RDS")
)

fm_joint_anno <- fm_joint
fm_joint_anno$pval <- z2p(fm_joint_anno$z)

fm_joint_anno <- anno_finemap_res(fm_joint_anno,
                                  snp_map = snp_map,
                                  mapping_table = mapping_table_joint,
                                  add_gene_annot = TRUE,
                                  map_by = "molecular_id",
                                  drop_unmapped = TRUE,
                                  add_position = TRUE,
                                  use_gene_pos = "mid")

top_regions_joint <- unique(fm_joint_anno$region_id[fm_joint_anno$group != "SNP" & fm_joint_anno$susie_pip > 0.8])
```

```{r joint-locusplots, fig.width=10, fig.height=8, results='asis'}
for (region_id in top_regions_joint) {
  cat(sprintf("\n#### Region: %s\n\n", region_id))

  print(make_locusplot(
    fm_joint_anno,
    region_id = region_id,
    ens_db = ens_db,
    weights = weights_joint,
    highlight_pip = 0.8,
    filter_protein_coding_genes = FALSE,
    filter_cs = F,
    color_pval_by = "cs",
    color_pip_by = "cs",
    panel.heights = c(4, 4, 1, 4)
  ))

  cat("\n\n")
}
```

### Peak Sets (caQTL in joint model)

A **peak set** is defined as all caQTL peaks that belong to the same credible set (CS) within the same LD region. The PIP of a peak set is the sum of the individual peak PIPs.

```{r joint-peak-sets}
peaks_in_cs_joint <- fm_joint[fm_joint$type == "caQTL" & !is.na(fm_joint$cs), ]

mapping_joint_caQTL <- readRDS("/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/ATAC_peak_mapping.RDS")
peaks_annot_joint <- merge(peaks_in_cs_joint, mapping_joint_caQTL, by = "molecular_id", all.x = TRUE)

peak_sets_joint <- do.call(rbind, lapply(
  split(peaks_annot_joint, list(peaks_annot_joint$region_id, peaks_annot_joint$cs), drop = TRUE),
  function(x) {
    region_parts <- strsplit(x$region_id[1], "_")[[1]]
    data.frame(
      region_chr   = as.integer(region_parts[1]),
      region_id    = x$region_id[1],
      cs           = x$cs[1],
      n_peaks      = nrow(x),
      sum_pip      = sum(x$susie_pip),
      peaks        = paste(x$molecular_id, collapse = ", "),
      contexts     = paste(x$context, collapse = ", "),
      peak_coords  = paste0("chr", x$chrom, ":", x$start, "-", x$end, collapse = "; "),
      stringsAsFactors = FALSE
    )
  }
))
peak_sets_joint <- peak_sets_joint[order(-peak_sets_joint$sum_pip), ]
rownames(peak_sets_joint) <- NULL

cat("**Total peak sets:**", nrow(peak_sets_joint), "\n")
```

### PIP Distribution of Peak Sets (joint)

```{r joint-peak-sets-pip-hist, fig.width=6, fig.height=4}
hist(peak_sets_joint$sum_pip, breaks = 20, col = "steelblue", border = "white",
     main = "Distribution of peak set PIPs (joint model)",
     xlab = "Sum PIP per peak set", ylab = "Count")
abline(v = 0.8, col = "red", lty = 2)
text(0.8, par("usr")[4] * 0.9, "PIP = 0.8", pos = 2, col = "red", cex = 0.8)
```

### High-Confidence Peak Sets (sum PIP > 0.8, joint)

```{r joint-peak-sets-high}
high_joint <- peak_sets_joint[peak_sets_joint$sum_pip > 0.8, ]

if (nrow(high_joint) > 0) {
  display_high_joint <- high_joint[, c("region_chr", "region_id", "cs", "n_peaks",
                                        "sum_pip", "peaks", "contexts", "peak_coords")]
  colnames(display_high_joint) <- c("Chr", "Region", "CS", "N peaks",
                                     "Sum PIP", "Peaks", "Contexts", "Peak coordinates")
  datatable(display_high_joint, rownames = FALSE,
            options = list(pageLength = 25, scrollX = TRUE),
            caption = "Peak sets with sum PIP > 0.8 (joint model)") %>%
    formatRound("Sum PIP", digits = 4)
} else {
  cat("No peak sets with sum PIP > 0.8.\n")
}
```

### Locus zoom plots for high-confidence peak sets (joint)

Locus zoom plots for regions containing peak sets with sum PIP > 0.8.

```{r joint-peakset-locusplot-prep}
top_regions_peakset_joint <- unique(high_joint$region_id)
```

```{r joint-peakset-locusplots, fig.width=10, fig.height=8, results='asis'}
for (region_id in top_regions_peakset_joint) {
  cat(sprintf("\n#### Region: %s\n\n", region_id))

  print(make_locusplot(
    fm_joint_anno,
    region_id = region_id,
    ens_db = ens_db,
    weights = weights_joint,
    highlight_pip = 0.8,
    filter_protein_coding_genes = FALSE,
    filter_cs = F,
    color_pval_by = "cs",
    color_pip_by = "cs",
    panel.heights = c(4, 4, 1, 4)
  ))

  cat("\n\n")
}
```
