---
title: "Preparing weights"
author: "XSun"
date: "2026-06-18"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

## Data source

QTL data from [Alasoo et al. 2018 (PMID 29379200)](https://pubmed.ncbi.nlm.nih.gov/29379200/), which mapped eQTLs and caQTLs (chromatin accessibility QTLs) in macrophages across 4 stimulation conditions: naive, SL1344, IFNg, and IFNg_SL1344.

Storaged in `/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200`

Input files: `{e,ca}QTL_{condition}_FastQTL_lead_only.txt.gz` — lead QTLs per gene/peak from FastQTL.

## How prediction weights were created

1. **Variant mapping**: rsIDs in the QTL files were mapped to b38 coordinates using the UKB LD reference variant info file (`/project2/compbio/UKB_LD_REF/ref_snp_info/ukb_b38_0.1_var_info.Rvar.gz`). This provides chrom, pos, ref, alt alleles for each rsID, used to construct `varID` in the format `chr{chrom}_{pos}_{ref}_{alt}_b38`.

2. **Weight creation**: Used `ctwas::create_predictdb_from_QTLs()` with:
   - `use_top_QTL = TRUE`: one QTL per gene/peak
   - `select_by = "weight"`: top QTL chosen by largest absolute effect size (slope)
   - `weight` = slope from FastQTL output
   - For eQTL files: `gene_table` includes gene name and gene_type = "protein_coding"
   - For caQTL files: no `gene_table` (peak IDs used directly)

3. **Output**: PredictDB format `.db` (SQLite) + `.txt.gz` (covariance) files in `weights/` directory.

4. **Mapping rate**: ~78% of caQTL SNPs and ~76% of eQTL SNPs were mapped to the UKB b38 reference. Unmapped SNPs include non-rsID variants (chr_pos format, ~1% of total) and rare variants absent from the UKB reference.

5. **Gene type annotation**: For eQTL weight files, gene biotype (e.g. protein_coding, ncRNA) was annotated using `org.Hs.eg.db` based on Ensembl gene IDs, rather than assuming all genes are protein-coding. Genes not found in the database are labeled as "unknown".


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
library(RSQLite)
library(data.table)
library(ggplot2)

weightdir <- "/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200/weights"
db_files <- list.files(weightdir, pattern = "\\.db$", full.names = TRUE)

read_db <- function(dbfile) {
  con <- dbConnect(SQLite(), dbfile)
  w <- as.data.table(dbReadTable(con, "weights"))
  e <- as.data.table(dbReadTable(con, "extra"))
  dbDisconnect(con)

  name <- gsub("\\.db$", "", basename(dbfile))
  parts <- strsplit(name, "_")[[1]]
  w[, qtl_type := parts[1]]
  w[, condition := paste(parts[-c(1, length(parts))], collapse = "_")]
  w[, file_name := name]

  w[, chr := as.integer(gsub("chr", "", sub("_.*", "", varID)))]

  if (nrow(e) > 0) {
    w <- merge(w, e[, .(gene, genename, gene_type)], by = "gene", all.x = TRUE)
  }
  w
}

all_weights <- rbindlist(lapply(db_files, read_db), fill = TRUE)
all_weights[, label := paste(qtl_type, condition, sep = " - ")]
```

## Overview

```{r summary_table}
summary_dt <- all_weights[, .(
  `Genes/Peaks` = uniqueN(gene),
  `Unique SNPs` = uniqueN(rsid),
  `Median |weight|` = round(median(abs(weight)), 4),
  `Max |weight|` = round(max(abs(weight)), 4)
), by = .(Type = qtl_type, Condition = condition)]

knitr::kable(summary_dt)
```

## Gene type breakdown (eQTL only)

```{r gene_type_table}
eqtl_data <- all_weights[qtl_type == "eQTL" & !is.na(gene_type)]
if (nrow(eqtl_data) > 0) {
  type_summary <- eqtl_data[, .(n = uniqueN(gene)), by = .(condition, gene_type)]
  type_wide <- dcast(type_summary, gene_type ~ condition, value.var = "n", fill = 0)
  knitr::kable(type_wide, caption = "Number of genes by biotype and condition")
}
```

```{r gene_type_barplot, fig.height=5}
if (nrow(eqtl_data) > 0) {
  type_counts <- eqtl_data[, .(n = uniqueN(gene)), by = .(label, gene_type)]
  ggplot(type_counts, aes(x = reorder(gene_type, -n), y = n, fill = gene_type)) +
    geom_col() +
    facet_wrap(~ label, ncol = 2) +
    labs(x = "Gene type", y = "Count", title = "Gene biotype distribution (eQTL)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
}
```

## Number of genes/peaks per weight file

```{r count_barplot, fig.height=5}
count_dt <- all_weights[, .(n = uniqueN(gene)), by = .(label, qtl_type)]

ggplot(count_dt, aes(x = reorder(label, n), y = n, fill = qtl_type)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(x = NULL, y = "Count", title = "Number of genes/peaks per weight file") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

## Distribution of weights

```{r weight_dist, fig.height=7}
ggplot(all_weights, aes(x = weight)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~ label, scales = "free_y", ncol = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Weight (slope)", y = "Count", title = "Distribution of weights") +
  theme_bw()
```

## Distribution of |weight|

```{r absweight_dist, fig.height=7}
ggplot(all_weights, aes(x = abs(weight))) +
  geom_histogram(bins = 50, fill = "darkorange", color = "white") +
  facet_wrap(~ label, scales = "free", ncol = 2) +
  labs(x = "|Weight|", y = "Count", title = "Distribution of absolute weights") +
  theme_bw()
```

## Chromosome distribution

```{r chr_dist, fig.height=7}
chr_counts <- all_weights[, .(n = .N), by = .(label, chr, qtl_type)]

ggplot(chr_counts, aes(x = factor(chr), y = n, fill = qtl_type)) +
  geom_col() +
  facet_wrap(~ label, scales = "free_y", ncol = 2) +
  labs(x = "Chromosome", y = "Count", title = "Genes/peaks per chromosome") +
  theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(size = 7))
```

## Overlap of genes/peaks across conditions

```{r overlap_eqtl, fig.height=5}
calc_overlap <- function(dt) {
  conditions <- unique(dt$condition)
  gene_lists <- lapply(conditions, function(c) unique(dt[condition == c, gene]))
  names(gene_lists) <- conditions

  n <- length(conditions)
  mat <- matrix(0, n, n, dimnames = list(conditions, conditions))
  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      mat[i, j] <- length(intersect(gene_lists[[i]], gene_lists[[j]]))
    }
  }
  mat
}

eqtl_overlap <- calc_overlap(all_weights[qtl_type == "eQTL"])
caqtl_overlap <- calc_overlap(all_weights[qtl_type == "caQTL"])

knitr::kable(eqtl_overlap, caption = "eQTL: gene overlap across conditions")
```

```{r overlap_caqtl}
knitr::kable(caqtl_overlap, caption = "caQTL: peak overlap across conditions")
```

## Weight comparison across conditions (eQTL)

```{r weight_compare_eqtl, fig.height=8}
eqtl_wide <- dcast(all_weights[qtl_type == "eQTL"], gene ~ condition, value.var = "weight")
conditions <- unique(all_weights[qtl_type == "eQTL", condition])

if (length(conditions) >= 2) {
  pairs <- combn(conditions, 2, simplify = FALSE)
  plots <- lapply(pairs, function(p) {
    tmp <- eqtl_wide[!is.na(get(p[1])) & !is.na(get(p[2]))]
    r <- round(cor(tmp[[p[1]]], tmp[[p[2]]], use = "complete.obs"), 3)
    ggplot(tmp, aes(x = .data[[p[1]]], y = .data[[p[2]]])) +
      geom_point(alpha = 0.3, size = 0.8) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
      labs(title = paste0(p[1], " vs ", p[2], "  (r=", r, ")")) +
      theme_bw()
  })
  gridExtra::grid.arrange(grobs = plots, ncol = 2)
}
```

## Weight comparison across conditions (caQTL)

```{r weight_compare_caqtl, fig.height=8}
caqtl_wide <- dcast(all_weights[qtl_type == "caQTL"], gene ~ condition, value.var = "weight")
conditions_ca <- unique(all_weights[qtl_type == "caQTL", condition])

if (length(conditions_ca) >= 2) {
  pairs <- combn(conditions_ca, 2, simplify = FALSE)
  plots <- lapply(pairs, function(p) {
    tmp <- caqtl_wide[!is.na(get(p[1])) & !is.na(get(p[2]))]
    r <- round(cor(tmp[[p[1]]], tmp[[p[2]]], use = "complete.obs"), 3)
    ggplot(tmp, aes(x = .data[[p[1]]], y = .data[[p[2]]])) +
      geom_point(alpha = 0.3, size = 0.8) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
      labs(title = paste0(p[1], " vs ", p[2], "  (r=", r, ")")) +
      theme_bw()
  })
  gridExtra::grid.arrange(grobs = plots, ncol = 2)
}
```
