---
title: "Preparing weights"
author: "XSun"
date: "2026-06-18"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Data source

QTL data from [Alasoo et al. 2018 (PMID 29379200)](https://pubmed.ncbi.nlm.nih.gov/29379200/), which mapped eQTLs and caQTLs in macrophages across 4 stimulation conditions: naive, SL1344, IFNg, and IFNg_SL1344.

Input files: `{e,ca}QTL_{condition}_RASQUAL_lead_only.txt.gz` — lead QTLs per gene/peak from RASQUAL.

# Multiple testing correction (from the paper)

The paper applies two layers of multiple testing correction:

1. **Step 1 — Correcting for multiple variants per gene (cis-window)**: Used EigenMT to estimate the number of independent tests in each gene's cis-window (±500kb). This gives:
   - `p_eigen`: nominal p-value corrected for the number of independent tests
   - `n_tests`: number of independent tests EigenMT estimated

2. **Step 2 — Correcting across genes**: Used Benjamini-Hochberg FDR on the EigenMT-corrected p-values. This gives:
   - `fdr_thresh`: the empirical FDR threshold — QTLs where `p_eigen < fdr_thresh` are significant

The paper reports "up to 3,431 genes (eQTLs) and 20,788 chromatin regions (caQTLs)" at 10% FDR. These numbers come from RASQUAL with `p_eigen < fdr_thresh`, which we reproduce exactly here.

# Data overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
library(RSQLite)
library(data.table)
library(ggplot2)

datadir <- "/project/xinhe/xsun/AD_jubao/data/QTL_pmid29379200"
rasqual_dir <- file.path(datadir, "RASQUAL")
```

```{r load_raw_data}
qtl_files <- list.files(rasqual_dir, pattern = "^(e|ca)QTL_.*_RASQUAL_lead_only\\.txt\\.gz$",
                        full.names = TRUE)

raw_list <- lapply(qtl_files, function(f) {
  qtl <- fread(f)
  fname <- basename(f)
  parts <- strsplit(gsub("_RASQUAL_lead_only\\.txt\\.gz$", "", fname), "_")[[1]]
  qtl[, qtl_type := parts[1]]
  qtl[, condition := paste(parts[-1], collapse = "_")]
  qtl[, label := paste(qtl_type, condition, sep = " - ")]
  qtl[, significant := p_eigen < fdr_thresh]
  qtl
})
raw_data <- rbindlist(raw_list, fill = TRUE)
```

```{r data_overview_table}
overview <- raw_data[, .(
  `Total genes/peaks` = .N,
  `Unique SNPs` = uniqueN(snp_id),
  `fdr_thresh` = round(fdr_thresh[1], 4),
  `Significant (p_eigen < fdr_thresh)` = sum(significant),
  `Median chisq` = round(median(chisq), 2),
  `Max chisq` = round(max(chisq), 2)
), by = .(Type = qtl_type, Condition = condition)]

knitr::kable(overview, caption = "Input RASQUAL QTL data overview")
```

# Creating prediction models

1. **Significance filtering**: Only genes/peaks with `p_eigen < fdr_thresh` were retained (matching the paper's method).

2. **Variant mapping**: rsIDs were mapped to b38 coordinates using the UKB LD reference variant info file (`/project2/compbio/UKB_LD_REF/ref_snp_info/ukb_b38_0.1_var_info.Rvar.gz`), constructing `varID` in the format `chr{chrom}_{pos}_{ref}_{alt}_b38`.

3. **Weight creation**: Used `ctwas::create_predictdb_from_QTLs()` with:
   - `use_top_QTL = TRUE`: one QTL per gene/peak
   - `select_by = "weight"`: since all weights are set to 1, one QTL is arbitrarily selected per gene/peak
   - `weight = 1` for all QTLs (effect size not informative when using top QTL only)


```{r load_weights}
read_db <- function(dbfile) {
  con <- dbConnect(SQLite(), dbfile)
  w <- as.data.table(dbReadTable(con, "weights"))
  e <- as.data.table(dbReadTable(con, "extra"))
  dbDisconnect(con)

  name <- gsub("\\.db$", "", basename(dbfile))
  parts <- strsplit(name, "_")[[1]]
  w[, qtl_type := parts[1]]
  w[, condition := paste(parts[-c(1, length(parts))], collapse = "_")]
  w[, file_name := name]
  w[, chr := as.integer(gsub("chr", "", sub("_.*", "", varID)))]

  if (nrow(e) > 0) {
    w <- merge(w, e, by = "gene", all.x = TRUE)
  }
  w
}

weightdir <- file.path(datadir, "RASQUAL", "weights_rasqual")
db_files <- list.files(weightdir, pattern = "\\.db$", full.names = TRUE)
all_weights <- rbindlist(lapply(db_files, read_db), fill = TRUE)
all_weights[, label := paste(qtl_type, condition, sep = " - ")]
```

## Summary of weight files

```{r summary_table}
summary_dt <- all_weights[, .(
  `Genes/Peaks` = uniqueN(gene),
  `Unique SNPs` = uniqueN(rsid)
), by = .(Type = qtl_type, Condition = condition)]

knitr::kable(summary_dt, caption = "Weight files summary (after p_eigen < fdr_thresh filtering and b38 mapping)")
```

## Number of genes/peaks per weight file

```{r count_barplot, fig.height=5}
count_dt <- all_weights[, .(n = uniqueN(gene)), by = .(label, qtl_type)]

ggplot(count_dt, aes(x = reorder(label, n), y = n, fill = qtl_type)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(x = NULL, y = "Count", title = "Number of genes/peaks per weight file") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

